Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-03-16 paper/arm.tex
Changed S4 to include M=v rather than [(Q /\ M=v)/Q]
Dealing with bad example 5.1.

@@

\begin{definition}
  \label{def:DS}
  % Let $[\FALSE/\D]$ be the substitution that performs $[\FALSE/\Dx{\aLoc}]$
  % for every $\aLoc$.
  Let $[\aForm/\Dx{*}]$ substitute $\aForm$ for every $\Dx{\bLoc}$.\\
  Let formula $\DL{\aLoc}{\amode}$ and substitution $\DS{\aLoc}{\amode}$ be defined:
  \begin{align*}
    \DL{\aLoc}{\mRLX}&=\TRUE
    &\DS{\aLoc}{\mRLX}&=[\TRUE/\Dx{\aLoc}] 
    \\
    \DL{\aLoc}{\mRA}&=\Dx{\aLoc}%\land\D{}
    &\DS{\aLoc}{\mRA}&=[\FALSE/\D]
    \\
    \DL{\aLoc}{\mSC}&=\Dx{\aLoc}%\land\D{}
    &\DS{\aLoc}{\mSC}&=[\FALSE/\D]
  \end{align*}
\end{definition}
\begin{definition}[\xCO/\xRASC/\xDGR]
  \label{def:pomsets-down}
  Update \refdef{def:pomsets-ra} to: % (\ref{L4} unchanged):
  \begin{enumerate}
  \item[\ref{S4})]
    $\aTr{\bEvs}{\bForm}$ implies
    \begin{math}
      \bForm
      \DS{\aLoc}{\amode}
      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}],
    \end{math}
  \item[\ref{S5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      \bForm
      \DS{\aLoc}{\amode}
      [\FALSE/\QS{\aLoc}{\amode}],
    \end{math}
  \item[\ref{L4})]
    $\aTr{\bEvs}{\bForm}$ implies
    \begin{math}
      \aVal{=}\aReg\limplies\bForm,
    \end{math}
  \item[\ref{L5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      \DL{\aLoc}{\amode}\land \bForm[\FALSE/\QL{\aLoc}{\amode}].
    \end{math}
  \end{enumerate}
\end{definition}
Load actions that require downgrading introduce $\Dx{\aLoc}$.
Relaxed stores on $\aLoc$ substitute true for $\Dx{\aLoc}$, whereas
synchronizing stores substitute false for $\Dx{\aLoc}$.

\begin{example}
  Revisiting \refex{ex:dgr1} and eliding irrelevant transformers:
  \begin{align*}
    \begin{gathered}[t]
      \PW{x}{2}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{a}{\DW{x}{2}}{}
          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of a}
          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
          \xo{a}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \PR[\mRA]{x}{r}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \raevent{b}{\DR[\mRA]{x}{2}}{}
          % \xform{x1d}{2{=}r\limplies\bForm}{below=of b}
          \xform{x1i}{\Dx{x} \land\bForm[\FALSE/\Qr{x}]}{below=of b} %x1d}
          % \xo{b}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \PW{y}{1}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{c}{\DW{y}{1}}{}
          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
          % \xo{c}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Associating right:
  \begin{align*}
    \begin{gathered}[t]
      \PW{x}{2}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{a}{\DW{x}{2}}{}
          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{right=.7em of a}
          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
          \xos{a}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \PR[\mRA]{x}{r}
      \SEMI
      \PW{y}{1}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \raevent{b}{\DR[\mRA]{x}{2}}{}
          \event{c}{\Dx{x}\mid\DW{y}{1}}{right=.7em of b}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Composing, we have, as desired:
  \begin{align*}
    \begin{gathered}[t]
      \PW{x}{2}
      \SEMI
      \PR[\mRA]{x}{r}
      \SEMI
      \PW{y}{1}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{a}{\DW{x}{2}}{}
          \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
          \event{c}{\DW{y}{1}}{right=of b}
          \rf{a}{b}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
\end{example}

@@

\begin{definition}[\xCO/\xRASC/\xDGR/\xRRD]
  \label{def:pomsets-arm}
  Let $[\aForm/\Q{}]$ be the substitution $[\aForm/\Qr{*}][\aForm/\Qw{*}][\aForm/\Qsc]$.  
  Update \refdef{def:pomsets-trans} to: % (\ref{L4} unchanged):
  \begin{enumerate}
  \item[\ref{S3})]
    $\labelingForm(\aEv)$ implies $\QS{\aLoc}{\amode}\land\aExp{=}\aVal$,
  \item[\ref{L3})]
    $\labelingForm(\aEv)$ implies $\QL{\aLoc}{\amode}$.
  % \item[\ref{T3})]
  %   $\labelingForm(\aEv)$ implies
  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
  %   \\
  %   $\labelingForm(\aEv)$ implies
  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
  \end{enumerate}
  \begin{enumerate}
  \item[\ref{S4})]
    $\aTr{\bEvs}{\bForm}$ implies
    \begin{math}
      %(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land
      \bForm %\noSUB{[\aExp/\aLoc]},
      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]
      \DS{\aLoc}{\amode}
    \end{math}
  \item[\ref{S5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      %\lnot\Qw{\aLoc}\land 
      \bForm %\noSUB{[\aExp/\aLoc]}.
      [\FALSE/\QS{\aLoc}{\amode}]
      \DS{\aLoc}{\amode}
    \end{math}
  \item[\ref{L4})]
    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
  \item[\ref{L5})]
    % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
    % \item[\ref{L6})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      % \lnot\Qx{\aLoc} \land
      \DL{\aLoc}{\amode}{} \land ((\aVal{=}\aReg\lor\RW)\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
    \end{math}
  \end{enumerate}  
\end{definition}

>>>>

\begin{example}
  Bad example:
  \begin{gather*}
    \PR{x}{r}\SEMI 
    \PW{x}{2}\SEMI
    \PR{x}{s}\SEMI
    \PW{y}{s{-}1} \PAR
    \PR{y}{r}\SEMI
    \PW{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{a}{\DR{x}{1}}{}
        \event{b}{\DW{x}{2}}{right=of a}
        \event{c}{\DR{x}{2}}{right=of b}
        \event{d}{\DW{y}{1}}{right=of c}
        \event{e}{\DR{y}{1}}{right=of d}
        \event{f}{\DW{x}{1}}{right=of e}
        \wk{a}{b}
        \rf{b}{c}
        \po{c}{d}
        \rf{d}{e}
        \po{e}{f}
        \rf[out=-165,in=-15]{f}{a}
      \end{tikzinline}}
  \end{gather*}
\end{example}

\begin{example}
  Should not wipe out local state only when there is a release-acquire pair,
  as in the definition of \jjr{}, which allows the following execution.  Even
  though there is a blocker $\DWP{x}{3}$, here we are using $\DWP{x}{0}$ to
  justify the lack of dependence $\DRP{x}{4}\xpo\DWP{w}{4}$.
  \begin{gather*}
    \PW{x}{0}\SEMI
    \IF{x}\THEN\PW{y}{2}\FI\SEMI
    \IF{\PR[\mRA]{z}{}}\THEN
      \PR{x}{s}\SEMI
      \IF{s\;\mathsf{even}}\THEN\PW{w}{4}\FI
    \FI
    \PAR
    \PW{x}{2}\SEMI
    \IF{\PR[\mRA]{y}{}}\THEN
      \PW{x}{3}\SEMI
      \PW[\mRA]{z}{1}
    \FI
    \SEMI
    \PW{x}{\PR{w}{}}
    \\
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{2a}{\DW{x}{0}}{}
        \event{2b}{\DR{x}{2}}{right=of 2a}
        \event{2c}{\DW{y}{1}}{right=of 2b}
        \raevent{2d}{\DR[\mRA]{z}{1}}{right=of 2c}
        \event{2e}{\DR{x}{4}}{right=of 2d}
        \event{2f}{\DW{w}{4}}{right=of 2e}
        %
        \event{3a}{\DW{x}{2}}{below=of 2a}
        \raevent{3b}{\DR[\mRA]{y}{1}}{right=of 3a}
        \event{3c}{\DW{x}{3}}{right=of 3b}
        \raevent{3d}{\DW[\mRA]{z}{1}}{right=of 3c}
        \event{3e}{\DR{w}{4}}{right=of 3d}
        \event{3f}{\DW{x}{4}}{right=of 3e}
        \wk{2a}{2b}
        \po{2b}{2c}
        \sync{2d}{2e}
        \sync[out=20,in=160]{2d}{2f}
        \sync{3b}{3c}
        \sync{3c}{3d}
        \po{3e}{3f}
        \rf{3a}{2b}
        \rf{2c}{3b}
        \rf{3d}{2d}
        \rf{3f}{2e}
        \rf{2f}{3e}
        \wk{2a}{3a}
        %\wk[out=-20,in=-160]{3c}{3f}
        %\wk[out=20,in=160]{2b}{2e}
      \end{tikzinline}}
  \end{gather*}
  (The read-clobbering semantics from \textsection\ref{sec:substitutions}
  does not change this.)
\end{example}

It does seem to me like a non-local blocker for program-order must include a
read of $x$ or a release (to get order from the write) and an acquire (to get
the order to the read).

What if an acquire preceded by a release or non-local read on $x$ wiped the
local state of $x$?  The problem with this kind of solution (and P6 in ooplsa
specifically) is that it destroys roach-motel.

@@

\begin{definition}
  \label{def:DS}
  % Let $[\FALSE/\D]$ be the substitution that performs $[\FALSE/\Dx{\aLoc}]$
  % for every $\aLoc$.
  Let $[\aForm/\Dx{*}]$ substitute $\aForm$ for every $\Dx{\bLoc}$.\\
  Let formula $\DL{\aLoc}{\amode}$ and substitution $\DS{\aLoc}{\amode}$ be defined:
  \begin{align*}
    \DL{\aLoc}{\amode}&=
    \begin{cases}
      \Dx{\aLoc} & \textif \amode\neq\mRLX
      \\
      \TRUE& \textotherwise
    \end{cases}
    &
    \DS{\aLoc}{\amode}&=
    \begin{cases}
      [\TRUE/\Dx{\aLoc}]  & \textif \amode=\mRLX
      \\
      [\TRUE/\Dx{\aLoc}][\FALSE/\D] & \textotherwise
    \end{cases}
  \end{align*}
  % \begin{align*}
  %   \DS{\aLoc}{\mRLX}&=[\TRUE/\Dx{\aLoc}] 
  %   &\DL{\aLoc}{\mRLX}&=\TRUE
  %   \\
  %   \DS{\aLoc}{\mRA}&=[\FALSE/\D]
  %   &\DL{\aLoc}{\mRA}&=\Dx{\aLoc}%\land\D{}
  %   \\
  %   \DS{\aLoc}{\mSC}&=[\FALSE/\D]
  %   &\DL{\aLoc}{\mSC}&=\Dx{\aLoc}%\land\D{}
  % \end{align*}
\end{definition}
\begin{definition}[\xCO/\xRASC/\xDGR]
  \label{def:pomsets-down}
  Update \refdef{def:pomsets-ra} to: % (\ref{L4} unchanged):
  \begin{enumerate}
  \item[\ref{S4})]
    $\aTr{\bEvs}{\bForm}$ implies
    \begin{math}
      \bForm
      \DS{\aLoc}{\amode} \land\aExp{=}\aVal,
      %[(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}],
    \end{math}
  \item[\ref{S5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      \bForm
      \DS{\aLoc}{\amode}
      [\FALSE/\QS{\aLoc}{\amode}],
    \end{math}
  % \item[\ref{L4})]
  %   $\aTr{\bEvs}{\bForm}$ implies
  %   \begin{math}
  %     \aVal{=}\aReg\limplies\bForm,
  %   \end{math}
  \item[\ref{L5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      \DL{\aLoc}{\amode}\land \bForm[\FALSE/\QL{\aLoc}{\amode}].
    \end{math}
  \end{enumerate}
\end{definition}
Load actions that require downgrading introduce $\Dx{\aLoc}$.
Relaxed stores on $\aLoc$ substitute true for $\Dx{\aLoc}$, whereas
synchronizing stores substitute false for $\Dx{\aLoc}$.

\begin{example}
  \label{ex:dgr2}
  Revisiting \refex{ex:dgr1} and eliding irrelevant transformers:
  \begin{align*}
    \begin{gathered}[t]
      \PW{x}{2}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{a}{\DW{x}{2}}{}
          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of a}
          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
          \xo{a}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \PR[\mRA]{x}{r}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \raevent{b}{\DR[\mRA]{x}{2}}{}
          % \xform{x1d}{2{=}r\limplies\bForm}{below=of b}
          \xform{x1i}{\Dx{x} \land\bForm[\FALSE/\Qr{x}]}{below=of b} %x1d}
          % \xo{b}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \PW{y}{1}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{c}{\DW{y}{1}}{}
          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
          % \xo{c}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Associating right:
  \begin{align*}
    \begin{gathered}[t]
      \PW{x}{2}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{a}{\DW{x}{2}}{}
          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{right=.7em of a}
          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
          \xos{a}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \PR[\mRA]{x}{r}
      \SEMI
      \PW{y}{1}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \raevent{b}{\DR[\mRA]{x}{2}}{}
          \event{c}{\Dx{x}\mid\DW{y}{1}}{right=.7em of b}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Composing, we have, as desired:
  \begin{align*}
    \begin{gathered}[t]
      \PW{x}{2}
      \SEMI
      \PR[\mRA]{x}{r}
      \SEMI
      \PW{y}{1}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{a}{\DW{x}{2}}{}
          \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
          \event{c}{\DW{y}{1}}{right=of b}
          \rf{a}{b}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
\end{example}

@@

\begin{definition}[\xCO/\xRASC/\xDGR/\xRRD]
  \label{def:pomsets-arm}
  Let $[\aForm/\Q{}]$ be the substitution $[\aForm/\Qr{*}][\aForm/\Qw{*}][\aForm/\Qsc]$.  
  Update \refdef{def:pomsets-trans} to: % (\ref{L4} unchanged):
  \begin{enumerate}
  \item[\ref{S3})]
    $\labelingForm(\aEv)$ implies $\QS{\aLoc}{\amode}\land\aExp{=}\aVal$,
  \item[\ref{L3})]
    $\labelingForm(\aEv)$ implies $\QL{\aLoc}{\amode}$.
  % \item[\ref{T3})]
  %   $\labelingForm(\aEv)$ implies
  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
  %   \\
  %   $\labelingForm(\aEv)$ implies
  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
  \end{enumerate}
  \begin{enumerate}
  \item[\ref{S4})]
    $\aTr{\bEvs}{\bForm}$ implies
    \begin{math}
      %(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land
      \bForm 
      \DS{\aLoc}{\amode}
      \land\aExp{=}\aVal %\noSUB{[\aExp/\aLoc]},
    \end{math}
  \item[\ref{S5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      %\lnot\Qw{\aLoc}\land 
      \bForm %\noSUB{[\aExp/\aLoc]}.
      \DS{\aLoc}{\amode}
      [\FALSE/\QS{\aLoc}{\amode}]
    \end{math}
  \item[\ref{L4})]
    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
  \item[\ref{L5})]
    % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
    % \item[\ref{L6})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      % \lnot\Qx{\aLoc} \land
      \DL{\aLoc}{\amode}{} \land
      ((\aVal{=}\aReg\lor\RW)\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]),
      %\textwhen \aEvs\neq\emptyset
    \end{math}
  \item[\ref{L6})] 
    $\aTr{\dEvs}{\bForm}$ implies
    \begin{math}
      \DL{\aLoc}{\amode}{} \land
      \bForm,
    \end{math}
    when $\aEvs=\emptyset$.
  \end{enumerate}  
\end{definition}
